---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Camber Vincent"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
getwd() #getting working directory

#installing needed packages
library(tidyverse)
library(agricolae)
library(lubridate)
library(here)

lake_chemphys<-read_csv("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",
                        show_col_types=F) #quiet warning message
lake_chemphys$sampledate<-mdy(lake_chemphys$sampledate) #convert date column

#2
theme_standard<-theme(
  
  text=element_text(family="Times",size=12,color="black"), #setting base font to Times New Roman
  plot.title=element_text(family="Helvetica",face="bold",size=16,  #title text theme
                          margin=margin(b=3)), #added margin for visual clarity
  plot.subtitle=element_text(family="Helvetica",face="italic",size=12, #subtitle text theme
                             color="gray20",
                             margin=margin(b=10)), #added margin for visual clarity
  
  plot.background=element_rect(fill="white"), #background set to white
  panel.background=element_rect(fill="white"), #graph background set to white
  panel.border=element_rect(color="black",fill=NA), #set a border around the graph

  panel.grid.major=element_line(color="gray85"), #recolor gridlines
  panel.grid.minor=element_line(color="gray95"),
  axis.ticks=element_blank() #turn off ticks
  
)

theme_set(theme_standard) #set custom theme as the base theme for all graphs
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no significant change in mean lake temperature recorded during July across all lakes.
Ha: There is a signifianct change in mean lake temperature recorded during July across all lakes.

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot,warning=FALSE}
#4
lake_chemphys_july<-lake_chemphys%>% #initiate pipe
  filter(format(sampledate,"%m")=="07")%>% #filtering using the sampledate column for July
  select(lakename,year4,daynum,depth,temperature_C)%>% #selecting columns
  drop_na() #removes the NA values

#5
ggplot(lake_chemphys_july,aes(x=depth,y=temperature_C))+
  geom_jitter(alpha=0.2)+ #jitter used and points made partially transparent for visual clarity
  geom_smooth(method="lm",color="red")+ #line color changed for clarity
  scale_x_continuous(limits=c(0,NA),expand=c(0,0))+ #calling expand to justify 0,0 coordinate
  scale_y_continuous(limits=c(0,35),expand=c(0,0))+ #setting y-limits
  labs(title="Temperature by Depth",
       subtitle="For All Lakes in July",
       x="Depth (m)",
       y="Temperature (Cº)") #adding labels

```

6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that as depth increaes, temperature will decrease. The distribution of points does not appear to be truly linear. The bulk of the data follows an reverse-S curve and appears to hit an asymptote around 5ºC.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
tempdepth.regression<-lm( #setting linear regression
  data=lake_chemphys_july,
  temperature_C~depth)

summary(tempdepth.regression) #displaying results
```

8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The intercept of the model is at 21.96ºC, suggesting that at a depth of 0 meters the average temperature is predicted to be 21.96ºC. The coefficient for depth is -1.95 which suggests that for every 1-meter increase in depth, temperature will decrease by approximately 1.95ºC. Both of these values (the intercept and slope) have a p-value less than 2e-16 and are therefore highlighy statistically signifanct at the p<0.001 level. The model has an R-squared value of 0.7387 which suggests that 73.87% of the variance in temperature can be explained by variance in depth. The model rests on 9726 degrees of freedom with a residual standard error of 3.835, meaning that the observed temperature deviates from the model's prediction by an average of 3.835ºC.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9



#10


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

```

>Answer: 
